---
title: "Untitled"
description: |
  A new article created using the Distill format.
author:
  - name: Nora Jones 
    url: https://example.com/norajones
    affiliation: Spacely Sprockets
    affiliation_url: https://example.com/spacelysprokets
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Much of the US is in a housing crisis, and Madison is no exception. Our population is growing, and the growth of housing stock has not kept pace, leading to higher rents and real estate prices. There have been iniatives to increase the production of housing, but how successful have they been? Inspired by a thread on Twitter, this post looks at building permit data from SOCDS.

```{r}
library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(tidycensus)
library(forcats)
```

# Population growth
Madison has been and continues to be one of the fasted growing cities in the state of Wisconsin. Between 1970 and 2020, the city added almost 100,000 new residents. That is an increase of about 56 percent.[^1]

[^1]: Some part of the growth is the result of expanding city limits; however, in general the newly annexed areas did not have large populations.]




```{r}
# Population time series Madison city
df <- read_csv("data/nhgis0002_ts_nominal_place.csv")


df %>% 
  filter(NHGISCODE == "G55048000" & YEAR %in% c("1970", "1980", "1990", "2000", "2010", "2020")) %>% 
  # mutate(difference = AV0AA - lag(AV0AA)) %>% 
  # relocate(difference)
  ggplot(aes(as.numeric(YEAR), AV0AA)) +
  geom_line() +
  xlab("Year") +
  ylab("Population")+
  scale_y_continuous(labels = scales::comma, limits = c(0, 275000)) +
  theme(panel.grid.minor.x = element_blank()) +
  labs(caption = "Data: US Census, IPUMS NHGIS, University of Minnesota, www.nhgis.org\nVisualization: Harald Kliems",
       title = "Madison's population grew from\n173,000 in 1970 to 270,000 in 2020 (+56%)") +
  hrbrthemes::theme_ipsum_rc()
```

These new residents all need housing, and in general household sizes have been shrinking, requiring even more housing units. What does housing production look like in the city and the surrounding metro area?

The Office of Policy Development
and Research within the Department of Housing and Urban Development (HUD) keeps track of building permits issues on a monthly basis. Monthly data is choppy: A single large development can easily throw off the total in a month. Thus we look at the data in half year periods:

```{r load-permits}
# housing permits Madison city
permits <- read_csv("data/BuildingPermits(1).csv")

#housing permits for Madison CBSA
CBSA <- read_csv("data/BuildingPermits - Madison CBSA.csv")

# combine and prep city and CBSA data
permits_all <- permits %>% 
  rbind(CBSA) %>% 
    filter(Location == "Madison, WI" | Location == "MADISON") %>% 
    mutate(date = ym(paste(Year, Month)),
         quarter = quarter(date),
         semester = semester(date),
         place = case_when(Location == "MADISON" ~ "City of Madison",
                              Location == "Madison, WI" ~ "Madison metro area")) %>% 
  filter(date < ymd("2022-07-01"))  #remove 0 values for future months

  


permits_all %>% 
  filter(`Series Code` == 1) %>% #total units
  group_by(Year, semester, place) %>% 
  summarise(permits = sum(Permits), date) %>% 
  distinct(Year, semester, permits, place, .keep_all = T) %>% 
  pivot_wider(names_from = place, values_from = permits) %>% 
  mutate(metro_without_city = `Madison metro area` - `City of Madison`) %>% 
  pivot_longer(cols = c(metro_without_city, `City of Madison`, `Madison metro area`), names_to = "place", values_to = "permits") %>% 
  filter(place %in% c("metro_without_city", "City of Madison")) %>% 
  ggplot(aes(date, permits, color = place)) +
  geom_line(size = 1) +
  ylim(0,2000)+
  hrbrthemes::scale_color_ipsum(name = NULL, labels = c("City of Madison only", "Madison metro area,\nwithout City")) +
  hrbrthemes::theme_ipsum_rc() +
  theme(legend.position = "right") +
  labs(title = "New housing units permitted in the City of\n Madison and the surrounding metro area", subtitle = "Both the total number and the trend over time are similar between the two",
       caption = "Data: https://socds.huduser.gov/permits/\nVisualization: Harald Kliems\nMadison metro area encompasses Dane, Green, Iowa, and Columbia counties")


```
Housing production in the City of Madison was strong during the early aughts. And in the surrounding metro area it was even stronger. But once the financial crisis hit, it took almost a decade until housing construction returned to previous levels. We can also see the shock the early stage of the COVID-19 pandemic, and how that permitting backlog cleared over the course of 2020. 


# Types of housing
The previous section describes the growth in total units of housing. But were these newly permitted units single-family homes or parts of multi-family housing?

```{r}
permits_all %>% 
  filter(`Series Code` <= 2) %>% 
  pivot_wider(names_from = Series, values_from = Permits, id_cols = c(Location, Year, Month)) %>% 
  janitor::clean_names() %>% 
  mutate(location = case_when(location == "Madison, WI" ~ "madison_metro",
                              location == "MADISON" ~ "madison_city")) %>% 
  group_by(year, location) %>% 
  summarise(total_units = sum(total_units), sf_units = sum(units_in_single_family_structures)) %>% 
  pivot_wider(names_from = location, values_from = c(total_units, sf_units)) %>% 
  mutate(metro_without_city_sf = sf_units_madison_metro - sf_units_madison_city,
         metro_without_city_total = total_units_madison_metro - total_units_madison_city,
         prop_sf_madison_city = sf_units_madison_city/total_units_madison_city,
         prop_sf_metro_without_city = metro_without_city_sf/metro_without_city_total) %>% 
  select(year, starts_with("prop")) %>% 
  pivot_longer(cols = starts_with("prop"), 
               names_to = "place", 
               values_to = "prop_sf",
               names_prefix = "prop_sf_") %>% 
ggplot(aes(year, prop_sf, color = fct_reorder(place, -prop_sf))) +
  geom_line(size = 2) +
  labs(title = "Proportion of newly permitted single family homes\nto total units permitted",
       subtitle = "The share of SF homes to total units permitted has declined over time. In the City of Madison the\nshare is much lower than in the surrounding metro area.",
       caption = "Data: https://socds.huduser.gov/permits/\nVisualization: Harald Kliems",
       x = "year",
       y = "Proportion single-family") +
  hrbrthemes::scale_color_ipsum(name = NULL, 
                       labels = c("Madison metro area,\nwithout City", "City of Madison only")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), 
                     limits = c(0,1)) +
  hrbrthemes::theme_ipsum_rc()
  
```



```{r acs-housing-units}
housing_units <- read_csv("data/nhgis0003_ts_nominal_place.csv")

housing_units %>% 
    filter(NHGISCODE == "G55048000" & YEAR %in% c("1970", "1980", "1990", "2000", "2010", "2020")) %>% 
  relocate(A41AA) %>% 
  ggplot(aes(as.numeric(YEAR), A41AA)) +
  geom_line() +
  xlab("Year") +
  ylab("Housing units")+
  scale_y_continuous(labels = scales::comma, limits = c(0, 140000)) +
  theme(panel.grid.minor.x = element_blank()) +
  labs(caption = "Data: US Census, IPUMS NHGIS, University of Minnesota, www.nhgis.org\nVisualization: Harald Kliems",
       title = "The number of housing units in Madison grew from\n57,000 in 1970 to 126,000 in 2020 (+121%)") +
  hrbrthemes::theme_ipsum_rc()

```


